% chap00_setup.tex - Installation and first steps
% ============================================================================

\chapter{Setting Up Your Environment}
\label{chap:setup}

Before writing any code, you need a working Python installation. This chapter has two parts: the essentials you need to start (Sections~\ref{sec:why-program} through~\ref{sec:venv}), and additional information, tips, tools, and practices for those interested (everything after).

To jump straight into Chapter~\ref{chap:basics}, you only need to complete the essential sections. They should take 15--30 minutes and will give you everything needed to learn to use Python.

% ----------------------------------------------------------------------------
\section{Why Scientists Program}
\label{sec:why-program}

You already know how to analyse data. You've used Excel, maybe Origin or Igor Pro. You can make graphs and calculate statistics. So why learn to program?

\textbf{Reproducibility.} When you analyse data by clicking through menus, the process lives only in your memory. Imagine months or even years later, someone asks how you normalised some data, which for some reason was a process everyone missed as missing in your methods. Good luck remembering which buttons you clicked, or perhaps even what software was used. A \emph{script}---a text file containing instructions for the computer---documents every step explicitly. Run it again and you get exactly the same result. You are essentially documenting your methods AND doing your analysis at the same time, while getting early warning that you forgot to document something, because your outputs wont be correct. This saves you a lot of time and effort.

\textbf{Automation.} You have 200 data files. Processing these one by one manually takes hours, weeks or even years. A script can processes them in seconds---and doesn't make mistakes at file 187 because it's tired. Say you finished half your files, only to realise you missed a step half-way through the process, now you have to start again. With a script you simply update the methodological error, and re-run.

\textbf{Flexibility.} Commercial software does what its developers anticipated consumers need. Your analysis might need something unusual: a custom baseline correction, a niche fitting function, or a way to merge data from three different instruments. With programming, you build exactly what you need.

\textbf{Figures.} Publication-quality figures require control over every detail: axis labels, font sizes, colour schemes, subplot layouts. Programming gives you that control. Once you write the code for a figure style you like, every future paper you write can use the same polished look.

\begin{labexample}
Imagine you collect UV-Vis spectra weekly to monitor a reaction. Each measurement produces a file. By the end of the project, you have 50 files. A Python script can:
\begin{itemize}
    \item Read all 50 files automatically
    \item Subtract baseline from each spectrum
    \item Calculate the absorbance at your wavelength of interest
    \item Plot the results vs. time
    \item Export the processed data and figure
\end{itemize}
When asked ``What if we tried this novel baseline method?''---you simply change your baseline correction line/s and re-run. Now investigating promising alternatives is no longer locked behind an effort threshold. Or, say you have a script that does what is required, you can get your analysis as the data is being produced, giving you early insights and feedback. Maybe this lets you terminate a set of experiments early, or warns you something is wrong with conditions, before you waste time and resources on more useless data. The list of benefits goes on and on, your creativity being the only real limit.
\end{labexample}

\section{Paths}
\label{sec:paths}

To tell Python where to find your data, you need to provide a \emph{path}---the digital address of a file or folder (also called a "directory") on your computer. 

\begin{itemize}
    \item \textbf{Absolute path:} The full address starting from the very "root" (for example, your hard drive). 
    \begin{itemize}
        \item Windows: \filepath{C:\textbackslash Users\textbackslash Alice\textbackslash Documents\textbackslash data.txt}
        \item macOS/Linux: \filepath{/home/alice/Documents/data.txt}
    \end{itemize}
    \item \textbf{Relative path:} The address relative to where you are \emph{right now} (your "Working Directory").
    \begin{itemize}
        \item \textbf{Windows} (uses backslashes \filepath{\textbackslash}):
            \begin{itemize}
            \item \filepath{config.txt} (In your current folder)
            \item \filepath{data\textbackslash spectrum.csv} (In the sub-folder "data")
            \item \filepath{..\textbackslash results.csv} (\filepath{..} means "go up one folder")
            \end{itemize}
        \item \textbf{macOS/Linux} (uses forward slashes \filepath{/}):
            \begin{itemize}
            \item \filepath{config.txt} (In your current folder)
            \item \filepath{data/spectrum.csv} (In the sub-folder "data")
            \item \filepath{../results.csv} (\filepath{..} means "go up one folder")
            \end{itemize}
    \end{itemize}
\end{itemize}

Think of slashes as the boundaries between folders. Every time you double-click a folder in a file browser, you are essentially adding a slash and a folder name to the path. Because slashes are "boundary markers," you cannot use them when naming a file or folder.

\begin{note}
Absolute paths may feel "safer" because they always point to the same place. However, if you send your code to a colleague, an absolute path like \filepath{C:\textbackslash Users\textbackslash Fritz...} will break on \emph{their} computer because they (probably) aren't also "Fritz." Relative paths are preferred for projects so that code will run as expected when moved between computers.
\end{note}

\begin{osbox}
\textbf{The Slash Direction:} Windows is the only major system that uses the backslash (\filepath{\textbackslash}). All other systems, and the internet, use the forward slash (\filepath{/}). If you see \filepath{./} at the start of a path, the dot simply means "the current folder." 

\textbf{Note:} While I will use Linux-style forward slashes (\filepath{/}) throughout the remainder of this book, Windows users should mentally swap them for backslashes (\filepath{\textbackslash}) when typing commands into their terminal.
\end{osbox}

% ----------------------------------------------------------------------------
\section{Installing Python}
\label{sec:install-python}

Python is free. You download it, install it, and use it forever without paying anyone. Several distributions exist; we'll use the standard one from python.org.

\begin{osbox}
Installation differs slightly by operating system. Follow the section for your system, then continue to the next section about the terminal.
\end{osbox}

\subsection{Windows}

\begin{enumerate}
    \item Go to \url{https://www.python.org/downloads/}
    \item Click the big yellow ``Download Python'' button (for this book, get version 3.10 or later, though generally not the very latest)
    \item Run the installer
    \item \textbf{Important:} Check the box that says ``Add Python to PATH''
    \item Click ``Install Now''
    \item When finished, click ``Disable path length limit'' if that option appears
\end{enumerate}

\begin{note}
What is ``PATH''? When you type a command like \shell{python}, your computer needs to know where to find the Python program. The PATH is a list of folders the computer searches through. ``Add Python to PATH'' means: ``remember where Python is installed so I can run it from anywhere.'' Without this, you'd have to type the full location of Python every time.
\end{note}

To verify the installation:
\begin{enumerate}
    \item Open PowerShell (search for it in the Start menu)
    \item Type \shell{python --version} and press Enter
    \item You should see something like \shell{Python 3.13.4}, select at least version 3.10 or later.
\end{enumerate}

\begin{warning}
Windows sometimes has an old ``python'' command that opens the Microsoft Store. If \shell{python --version} opens the Store instead of showing a version number, try \shell{python3 --version} instead. The ``Add Python to PATH'' checkbox during installation usually prevents this.
\end{warning}

\subsection{macOS}

macOS comes with an old Python 2, which we don't want. Install Python 3:

\begin{enumerate}
    \item Go to \url{https://www.python.org/downloads/}
    \item Download the macOS installer (version 3.10 or later)
    \item Open the downloaded \filepath{.pkg} file
    \item Follow the installation wizard
\end{enumerate}

To verify:
\begin{enumerate}
    \item Open Terminal (find it in Applications > Utilities, or search with Spotlight)
    \item Type \shell{python3 --version} and press Enter
    \item You should see the version number
\end{enumerate}

\begin{note}
On macOS, always use \shell{python3} and \shell{pip3} instead of \shell{python} and \shell{pip}. The shorter commands often point to the old system Python 2.
\end{note}

\subsection{Linux}

Most Linux distributions include Python 3. Check first:

\begin{lstlisting}[style=shellstyle]
python3 --version
\end{lstlisting}

If Python 3.10 or later is installed, you're ready to continue. If not, or if you simply want a newer version:

\textbf{Ubuntu/Debian:}
\begin{lstlisting}[style=shellstyle]
apt update
apt install python3 python3-pip python3-venv
\end{lstlisting}

\textbf{Fedora:}
\begin{lstlisting}[style=shellstyle]
dnf install python3 python3-pip
\end{lstlisting}

\begin{note}
\textbf{Administrative Privileges Required}

These commands modify system files and require administrative access. You will likely need to prepend \py{sudo} (e.g., \py{sudo apt update}) to execute them.

I intentionally omit \py{sudo} from the examples. It is good practice to type \py{sudo} manually, as it forces a moment of pause to verify exactly what command you are about to run with root privileges. You should never blindly copy-paste shell commands---especially those involving superuser privileges---without first understanding what the command will do.
\end{note}

% ----------------------------------------------------------------------------
\section{The Terminal: Talking to Your Computer}
\label{sec:launcher}

You're used to interacting with your computer by clicking icons and menus. But there's another way: typing commands directly. The program where you type these commands is called a \emph{terminal} (also called a ``command line'' or ``shell''). Think of it as having a text conversation with your computer instead of pointing at things.

Why use the terminal? Because many programming tools---including Python---are designed to be run by typing commands. It's also faster once you learn the basics, and it's the universal way programmers work across all operating systems.

Each operating system has its own built in terminal program:

\begin{itemize}
    \item \textbf{Windows:} Command Prompt or PowerShell (both pre-installed; search in Start menu). PowerShell is the newer one.
    \item \textbf{macOS:} Terminal (in Applications > Utilities, or search with Spotlight)
    \item \textbf{Linux:} Terminal (varies by distribution; usually Ctrl+Alt+T opens it)
\end{itemize}

You'll also hear about Anaconda Prompt. Anaconda is an alternative Python distribution popular in data science. It's fine, but adds complexity. We'll use standard Python to keep things clean.

Open your terminal now. You'll see a \emph{prompt}---some text followed by a cursor, waiting for you to type something:

\begin{lstlisting}[style=shellstyle]
PS C:\Users\yourname>          # Windows PowerShell
C:\Users\yourname>             # Windows Command Prompt
yourname@computer ~ %          # macOS Terminal
yourname@computer:~$           # Linux Terminal
\end{lstlisting}

The exact appearance varies, but they all work the same way: you type a \emph{command} (an instruction) and press Enter. The computer does what you asked, shows you the result, and waits for the next command.

% ----------------------------------------------------------------------------
\section{Your First Python Session}
\label{sec:first-repl}

Let's verify Python is working. In your terminal, type:

\begin{lstlisting}[style=shellstyle]
python          # Windows
python3         # macOS/Linux
\end{lstlisting}

You should see something like:

\begin{lstlisting}[style=outputstyle]
Python 3.11.4 (main, Jun  9 2023, 07:30:55)
[GCC 12.2.0] on linux
Type "help", "copyright", "credits" or "license" for more information.
>>>
\end{lstlisting}

The \py{>>>} is Python's prompt. Try a quick calculation:

\begin{lstlisting}
>>> 2 + 2
4
>>> 7 * 6
42
\end{lstlisting}

If Python responds with the correct answers, your installation is working. To exit, type:

\begin{lstlisting}
>>> exit()
\end{lstlisting}

Or press Ctrl+D (macOS/Linux) or Ctrl+Z then Enter (Windows).

\begin{tip}
You've just used Python's interactive mode, often called the REPL (Read-Eval-Print Loop). It's useful for quick experiments and calculations. But don't worry about memorising this now---we'll cover interactive mode properly in Chapter~\ref{chap:basics}.
\end{tip}


% ----------------------------------------------------------------------------
\section{Virtual Environments}
\label{sec:venv}

This section might seem bureaucratic, but it solves real problems. Paying attention here will save you many headaches in the future, while making a lot of public code more accessible to you.

\subsection{The Problem}

Python by itself can do basic things, but its real power comes from \emph{packages} (also called \emph{libraries})---collections of code that other people have written and shared. Think of them like apps for your phone: Python is the operating system, and packages add new abilities.

While you \emph{could} write your own statistical analysis tools from scratch, it is far more efficient to leverage the vast ecosystem of tools made available for free. For example, \textbf{NumPy} handles numerical computing, \textbf{Matplotlib} makes plots, and \textbf{Pandas} manages data tables. A major strength of Python is this community ecosystem. You'll install and use many of them.

But here's the problem: different projects might need different versions of the same package. Project A needs a package called \py{NumPy} at version 1.24 because it has a new and improved function; Project B needs \py{NumPy} version 1.21 because it uses a deprecated function that isn't supported in the new version.

If you install packages globally (available everywhere on your computer), you can only have one version. Updating for Project A breaks Project B and vice-versa. This is called ``dependency hell,'' and it's why scientists often say ``I reinstalled Python and now nothing works.''

\subsection{The Solution}

Virtual environments. Each project gets its own isolated Python environment with its own packages. Project A's environment has \py{NumPy} 1.24. Project B's environment has \py{NumPy} 1.21. They don't interfere with each other.

There are several external tools that manage environments (e.g., Conda, Poetry, Pipenv). While I highly recommend looking into \py{uv} (by Astral) once you are comfortable, this book focuses on Python's standard, built-in tools. Mastering these first ensures you understand the core mechanics without needing to install extra software, and provides the foundation needed to effectively use power tools like \py{uv} later.

\subsection{Creating a Virtual Environment}

First, we need a place for your project to live. Instead of switching back to your file explorer and clicking around, we can do this directly from the terminal.

\begin{enumerate}
    \item \textbf{Navigate} to where you keep your projects (e.g., Documents):
    \begin{lstlisting}[style=shellstyle]
    cd Documents
    \end{lstlisting}

    \item \textbf{Create} a new folder using the \emph{make directory} command (\py{mkdir}):
    \begin{lstlisting}[style=shellstyle]
    mkdir learning
    \end{lstlisting}

    \item \textbf{Enter} the new folder:
    \begin{lstlisting}[style=shellstyle]
    cd learning
    \end{lstlisting}
\end{enumerate}

\begin{tip}
Don't type long folder names manually! Type the first few letters (e.g., \py{cd Doc}) and press \textbf{Tab}. The terminal will auto-complete the rest. This prevents typos and makes navigation significantly faster.
\end{tip}

Create a virtual environment (we'll call it \filepath{venv}):

\begin{lstlisting}[style=shellstyle]
python -m venv .venv         # Windows
python3 -m venv .venv        # macOS/Linux
\end{lstlisting}

This tells \py{python} to use a module (\py{-m}) called \py{venv} to create a virtual environment in the folder \filepath{.venv}. This now contains a private Python installation.

\subsection{Activating the Environment}

Before using the environment, you must activate it. Make sure your prompt shows the file-path your virtual environment is in then enter:

\begin{osbox}
\textbf{Windows (PowerShell):}
\begin{lstlisting}[style=shellstyle]
.\.venv\Scripts\Activate.ps1
\end{lstlisting}

\textbf{Windows (Command Prompt):}
\begin{lstlisting}[style=shellstyle]
.venv\Scripts\activate.bat
\end{lstlisting}

\textbf{macOS/Linux:}
\begin{lstlisting}[style=shellstyle]
source .venv/bin/activate
\end{lstlisting}
\end{osbox}

Remember to use the exact name you gave your virtual environment. If you have been following my suggestion, it should be called \filepath{.venv} not \filepath{venv}. It is fine if it is called \filepath{venv}, or even \filepath{XaswefasDgfasdP}---while I recommend clean and informative names, it is ultimately up to you. Virtual environments are specific to the machine they were created on, and so should not be copied or shared. Therefore your virtual environment naming conventions should only ever impact you.

When activated, your prompt should change to show the environment name. As an example, if you are using PowerShell on Windows, you will see something like this:

\begin{lstlisting}[style=shellstyle]
(venv) PS C:\Users\yourname\Documents\python_projects\learning>
\end{lstlisting}

That \shell{(venv)} prefix means the virtual environment is active. Any packages you install or code you execute will now use the packages and python version from this specific environment.

\begin{osbox}
Windows blocks PowerShell scripts by default. So if you get an error about executing scripts, you can fix this by either using the command prompt instead, or doing the following:
\begin{enumerate}
    \item Close PowerShell.
    \item Right-click the PowerShell icon and select \textbf{Run as Administrator}. 
    \item Click \textbf{Yes} on the pop-up.
    \item Type this command and press Enter:
\begin{lstlisting}[style=shellstyle]
Set-ExecutionPolicy -ExecutionPolicy RemoteSigned -Scope CurrentUser
\end{lstlisting}
    \item Type \shell{Y} when asked.
\end{enumerate}
You can now close this window and go back to your normal PowerShell. You only have to do this once!
\end{osbox}

\subsection{Installing Packages}

With the environment activated, use \shell{pip} to install packages. Pip is Python's package manager---a tool that downloads packages from the internet and installs them for you. Think of it like an app store that you access by typing commands.

\begin{lstlisting}[style=shellstyle]
pip install numpy matplotlib pandas
\end{lstlisting}

This downloads NumPy, Matplotlib, and Pandas from the Python Package Index (a central repository of Python packages) and installs them into your virtual environment. To see what's installed:

\begin{lstlisting}[style=shellstyle]
pip list
\end{lstlisting}

\subsection{Saving and Reproducing Environments}

After installing packages, save the list to a file:

\begin{lstlisting}[style=shellstyle]
pip freeze > requirements.txt
\end{lstlisting}

This creates a \filepath{requirements.txt} file listing every package and its exact version. When setting up the project on another computer (or helping a collaborator), they can recreate your environment:

\begin{lstlisting}[style=shellstyle]
pip install -r requirements.txt
\end{lstlisting}

\begin{labexample}
Your collaborator can't reproduce your analysis. ``It works on my machine,'' you say. They're missing a package, or have a different version. Send them your \filepath{requirements.txt} and tell them to:
\begin{lstlisting}[style=shellstyle]
python3 -m venv venv
source venv/bin/activate
pip install -r requirements.txt
\end{lstlisting}
Now they have exactly your environment.
\end{labexample}

\subsection{Deactivating}

When you're done working, deactivate the environment:

\begin{lstlisting}[style=shellstyle]
deactivate
\end{lstlisting}

The \shell{(venv)} prefix disappears. You're back to the system Python.

\begin{tip}
Always activate your project's virtual environment before working. Make it a habit, like putting on safety glasses before entering the lab. Many ``my code stopped working'' problems are just forgotten activation.
\end{tip}

% ============================================================================
% PART 2: ADDITIONAL RESOURCES, TOOLS, AND BEST PRACTICES
% ============================================================================

\section*{Part 2: Additional Resources, Tools, and Best Practices}

You've now completed the essential setup. Everything that follows covers optional tools, alternatives, and best practices that enhance your workflow. You can read these sections at your own pace, or skip them entirely if you're eager to start learning Python in Chapter~\ref{chap:basics}. You can always return to them later when you need them.

% ============================================================================
% PART 2 CONTENT BEGINS
% ============================================================================

\section{Computer Hygiene}
\label{sec:hygiene}

Before installing anything, let's establish some habits that prevent future headaches.

\subsection{Folder Organisation}

Create a dedicated folder for your programming work. Don't scatter scripts across your Desktop, Downloads, and random project folders. Something like:

\begin{lstlisting}[style=shellstyle]
Documents/
    python_projects/
        learning/           # exercises from this book
        lab_analysis/       # your research scripts
        utilities/          # small helper scripts
\end{lstlisting}

Within each project, keep a consistent structure:

\begin{lstlisting}[style=shellstyle]
titration_project/
    data/               # raw data files (never modify these)
    output/             # processed data, figures
    scripts/            # your Python code
    README.txt          # notes about what this project does
\end{lstlisting}

\begin{warning}
Never put spaces in folder or file names used with code. Use underscores instead: \filepath{titration\_data.csv}, not \filepath{titration data.csv}. Spaces cause problems with many programming tools.
\end{warning}

\subsection{File Naming}

Good file names are:
\begin{itemize}
    \item Descriptive: \filepath{analyze\_kinetics.py}, not \filepath{script1.py}
    \item Dated when relevant: \filepath{calibration\_2024\_03\_15.csv}
    \item Consistent: pick a convention and stick with it
\end{itemize}

\subsection{Backup}

Your code represents hours of work. Back it up. Options:

\begin{itemize}
    \item Cloud sync (Dropbox, OneDrive, Google Drive): automatic, easy
    \item Git: a system that tracks every change you make to files, letting you undo mistakes and see your project's history. Best for code, but has a steeper learning curve.
    \item External drive: simple but requires discipline
\end{itemize}

At minimum, use cloud sync for your \filepath{python\_projects} folder. Git is worth learning, but explaining its use is beyond the scope of this book.

% ----------------------------------------------------------------------------
\section{Three Ways to Run Python Code}
\label{sec:three-ways}

You've seen the REPL. It's great for quick experiments, but not for serious work. Let's cover all three ways to run Python code.

\subsection{Interactive REPL}

You already know this one. Type \shell{python} (or \shell{python3}), and you're in an interactive session. Good for:
\begin{itemize}
    \item Quick calculations
    \item Testing small code snippets
    \item Exploring how functions work
\end{itemize}

Not good for:
\begin{itemize}
    \item Anything longer than a few lines
    \item Code you want to save and reuse
\end{itemize}

\subsection{Python Scripts}

A script is a text file containing Python code, saved with a \filepath{.py} extension. You write the code once, then run it whenever you want.

Create a file called \filepath{hello.py} with this content:

\begin{lstlisting}
# hello.py - My first Python script
print("Hello from a script!")
print("This is more than one line.")
result = 6 * 7
print("Six times seven is", result)
\end{lstlisting}

\begin{note}
In the last line, we pass two items to \py{print()}, separated by a comma. Python automatically puts a space between them. This is a simple way to combine text with variable values. In Chapter~\ref{chap:basics}, you'll learn about \emph{f-strings}, a more powerful approach for formatting output.
\end{note}

Run it from the terminal:

\begin{lstlisting}[style=shellstyle]
python hello.py         # Windows
python3 hello.py        # macOS/Linux
\end{lstlisting}

Output:
\begin{lstlisting}[style=outputstyle]
Hello from a script!
This is more than one line.
Six times seven is 42
\end{lstlisting}

Scripts are how you'll write most of your code. They're saved, editable, and shareable.

\subsection{Jupyter Notebooks}

Jupyter notebooks mix code, output, and explanatory text in one document. They're popular for data exploration and creating reports.

Install Jupyter in your virtual environment:

\begin{lstlisting}[style=shellstyle]
pip install jupyter
\end{lstlisting}

Launch it:

\begin{lstlisting}[style=shellstyle]
jupyter notebook
\end{lstlisting}

A browser window opens with the Jupyter interface. You can create notebooks, write code in cells, and run them interactively.

Notebooks are excellent for:
\begin{itemize}
    \item Exploratory data analysis
    \item Creating reports with embedded figures
    \item Teaching and documentation
\end{itemize}

Less ideal for:
\begin{itemize}
    \item Large, complex programs
    \item Code that runs automatically (without human interaction)
    \item Version control with Git
\end{itemize}

We'll mostly use scripts in this book because they teach better programming habits. But know that notebooks exist---you'll encounter them in the wild.

% ----------------------------------------------------------------------------
\section{Choosing an Editor}
\label{sec:editor}

You need a text editor to write scripts. Not Microsoft Word---that's a word processor that adds invisible formatting codes. You need an editor that saves \emph{plain text}: exactly what you type, nothing more.

\subsection{Start Simple: Basic Text Editors}

The best way to start is with a simple text editor. This might seem primitive, but it teaches you something important: Python scripts are just text files. You don't need special software to write Python---any program that saves plain text will work. This matters because it means your code isn't trapped in any particular tool.

\begin{itemize}
    \item \textbf{Windows:} The built-in Notepad works, though Notepad++ is better (free download from \url{https://notepad-plus-plus.org/}). Notepad++ has ``syntax highlighting''---it colours different parts of your code differently, making it easier to read.
    \item \textbf{macOS:} TextEdit works if you set it to plain text mode (Format > Make Plain Text). BBEdit (free version at \url{https://www.barebones.com/products/bbedit/}) is friendlier.
    \item \textbf{Linux:} gedit, kate, or your distribution's default text editor.
\end{itemize}

\subsection{Upgrading Later: Integrated Development Environments}
\label{subsec:ide}

\begin{tip}
If you are totally new to Python, you can reduce the information overload by skipping this section for now---a good time to come back is after Chapter~\ref{chap:functions}. Don't worry, I'll remind you!

I promise you're not missing out; the opposite is true. I recommend starting with text editors and never giving them up entirely. IDEs bring a lot to the table, but they take a lot as well.

If you already know your way around an IDE and disagree, then consider this tip a ponder block. Begin by asking yourself if the friction of a no-frills text editor might actually help \textbf{weaponize your divine laziness}.
\end{tip}

An IDE (Integrated Development Environment) is an editor designed specifically for programming. It provides helpful features: syntax highlighting, automatic error detection, tools to explore variables while your code runs, and more. When you're ready to upgrade from a basic text editor, here are some popular options.

\textbf{Spyder:} Designed specifically for scientists and data analysts using Python. If you've used RStudio, Spyder will feel familiar---it has a similar layout panels to show your script, an interactive console, a variable explorer, help files, a plot panel, and so on.

You could download it from \url{https://www.spyder-ide.org/} and install it using familiar pathways. Alternatively, check out the install guide \url{https://docs.spyder-ide.org/current/installation.html} for options such as installing using the build in package manager \py{pip}.

\textbf{Pyzo:} Similarly to Spyder, it was designed specifically for scientists. The main difference is that its interface is less cramped and therefore easier to understand for beginners. It is also flexible. It can be downloaded from \url{https://pyzo.org/start.html}.

\textbf{Visual Studio Code (VS Code):} Made by Microsoft, free, and extremely popular in the broader software development community (though not required). It is a general-purpose programming editor used for many languages. Download from \url{https://code.visualstudio.com/}. After installing, you'll need to add the Python extension: click the Extensions icon (four squares) in the sidebar, search for ``Python,'' and install it.

\textbf{No IDE:} A plain text editor like \py{nano}, \py{micro}, \py{vim}, or Notepad++ is perfectly serviceable. In fact, many scientists eventually work on remote supercomputers or cloud servers where graphical interfaces (like Spyder) aren't available. The ultimate IDE is your own mind. So, if you are comfortable editing plain text, you can work anywhere---on a powerful cluster, a colleague's laptop, or even on paper while commuting. The power lies in your abstractions made code, not the tool used to write it down. Challenge yourself to text-editor-only projects from time to time.

\begin{note}
VS Code is a generalist tool, whereas Spyder is a specialist. VS Code requires significant manual configuration to match Spyder's out-of-the-box scientific features. I recommend holding off on VS Code until you encounter a specific need that Spyder or Pyzo cannot fulfil.
\end{note}

\begin{note}
If you don't have access to a proper computer at a given time, you might want to use IDE mobile apps. Their interface is not as easy to navigate, and coding with an on-screen keyboard is not the most efficient, but still more than writing it on paper, as you will be able to send the code directly to your computer later.
\end{note}

% ----------------------------------------------------------------------------
\section{Chapter Summary}
\label{sec:setup-summary}

You now have:

\begin{itemize}
    \item Python installed and verified
    \item A terminal where you can type commands to your computer
    \item Understanding of virtual environments and packages (and why they matter)
    \item Knowledge of three ways to run Python: interactive mode (REPL), scripts, and notebooks
    \item A text editor for writing code (and knowledge of Spyder for later)
\end{itemize}

Before moving on, make sure you can:

\begin{enumerate}
    \item Open your terminal and start Python
    \item Create and activate a virtual environment
    \item Install a package with pip
    \item Create a \filepath{.py} file and run it from the terminal
\end{enumerate}

If any of these don't work, debug now. Search for your error message online. The setup must work before you can learn programming.

\begin{ponder}
Why do we use virtual environments instead of installing packages globally? What problem would arise if two of your projects needed different versions of NumPy?
\end{ponder}
